Is this the longest Chess game?

### 2020-03-23 trunk/chess/papers/longest.tex
This year, not 5 years ago

% XXX search for common typo "movies"

@@

\title{Is this the longest Chess game?}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2020 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2015 with the {\tt ShortMoveString} of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
\textcurrency 17,697
}
}

@@

In my experience, most chess games end in a few moves. Players who
seek longevity typically do so by playing many games of chess, rather
than a protracted XXX

(shortest game is well known)

Although many people ``know how to play'' chess, almost nobody fully
understands the rules of chess, most authoritatively given by
FIDE~\cite{fiderules}. (See for example Figure~\ref{fig:castle960} for
a minor chess scandal that erupted in 2019 over an obscure corner case
in the rules.) Several of these ``deep cuts'' have to do with game-ending
conditions that were introduced to keep the game from going on forever.

@@

% This is good for the insurers and reinsurers of chess tournaments
% (games are guaranteed to end, rather than simply ending with
% probability 1).
So, chess is formally a finite game. This good for computer
scientists, since it means that chess has a trivial O(1) optimal
solution. This allows us to move onto other important questions, like:
What is the longest chess game? In this paper I show how to compute
such a game, and then gratuitously present all of its 17,697 
moves. Even if you are a chess expert (``chexpert''), I bet you
will be surprised at some of the corner cases in the rules that are
involved.

@@

\begin{figure}
  \begin{center}
    \chessboard[
      % fillarea=a3-h1,printarea=a3-h1,
      setfen=2nr1r1k/pb1n1pqp/1p2p1p1/1N1pN3/5P2/1P4P1/P2PPQ1P/2R1K1RB,
      pgfstyle=curvemove,markmoves={e1-g1},
      pgfstyle=straightmove,markmoves={g1-f1},
      showmover=false]
  \end{center}
  \caption{ ({\it Nepomniachtchi -- So, 2019.}) White to move during a
    speed Chess960 (aka ``Fischer Random'') tournament. In this
    variant, the pieces start in different positions, but castling
    rules are such that the king and rook end up on the same squares
    that they would in normal chess. As a result, it is possible for
    the king or rook to not move during castling, or for the
    destination square for the king to already be occupied by the
    rook. Attempting to castle in the position depicted, grandmaster
    Ian Nepomniachtchi first touched the rook to move it out of the
    way. However, piece-touching rules require that when castling, the
    player first moves the king (and ``Each move must be played with
    one hand only''); but how? The rook is occupying g1! One commenter
    suggested tossing the king into the air, then sliding the rook to
    f1 while the king is airborne, and then watching the king land
    dead center on its target. The arbiter required Nepo to make a
    rook move instead, but this was later appealed, and the game
    replayed. } \label{fig:castle960}
\end{figure}

@@

\begin{quote}
\begin{itemize}
  \item[9.2.2.] Positions are considered the same if and only if the
same player has the move, pieces of the same kind and colour occupy
the same squares and the possible moves of all the pieces of both
players are the same. Thus positions are not the same if:
\begin{itemize}
\item[9.2.2.1.]	at the start of the sequence a pawn could have been captured en passant
\item[9.2.2.2.]	a king had castling rights with a rook that has not been moved, but forfeited these after moving. The castling rights are lost only after the king or rook is moved.
\end{itemize}
\end{itemize}
\end{quote}

@@

% (I seem to remember that this rule used to be about moves, not positions,
% but someone discovered infinite sequences of moves that don't repeat, right?
% I can't find this citation. Maybe can recreate those sequences.)
%
% Fascinatingly (at least as fascinating as anything in this
% questionable undertaking), this rule

@@

\begin{figure}[htp]
  \begin{center}
    \chessboard[setfen=8/8/5K2/8/8/8/Q7/k7 b - - 0 1]
  \end{center}
  \caption{Black to move and draw in 0 (!). Most players and even
    chexperts believe that the only legal move is Kxa2, and
    that the game then ends in a draw with ``insufficient material.''
    However, this game is {\em already over}. Since neither black
    nor white can win via any series of legal moves, by rule 5.2.2
    the game immediately ends in a `dead position'. (Although see
    Section~\ref{sec:ambiguity} for possible ambiguity in this
    rule.)} \label{fig:insufficient}
\end{figure}

@@

\begin{figure}[htp]
  \begin{center}
    \chessboard[setfen=k1bB4/8/2p1p1p1/1pPpPpP1/1P1P1P2/8/8/K1Bb4 w - - 0 1]
  \end{center}
  \caption{Thinking of implementing the rules of chess? To be correct,
    you'll need your program to be able to deduce that no helpmates are
    possible in this position and thus the game is over. Stockfish rates
    this as +0.4 for white, even searching to depth 92.
    (Position is due to user {\tt supercat} on Chess StackExchange.)
  } \label{fig:nohelp}
\end{figure}

@@

First, the rules specifically define ``legal move'' in Article~3, and
3.10.1 says ``A move is {\em legal} when all the relevant requirements
of Articles 3.1 -- 3.9 have been fulfilled.'' These requirements
describe the movement of each piece as you are familiar (e.g. 3.3
``The rook may move to any square along the file or the rank on which
it stands.''). They also disallow capturing one's own pieces, or
moving when in check. However, they allow as legal some moves that
would otherwise be prohibited, like capturing the opponent's king
(this is excluded by 1.4.1, outside the definition of ``legal'').
Capturing the opponent's king is generally not useful for
demonstrating that checkmate is possible
(Figure~\ref{fig:kingcapture}), so this is mostly a curiosity.

\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \chessboard[smallboard,marginleft=false,marginright=false,setfen=BBKBRRnk/PPPPPPnn/3P1BNR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] &
      \chessboard[smallboard,marginright=false,setfen=BBKBRR1B/PPPPnP1n/3P2NR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] \\
      (a) & (b) \\
  \end{tabular}
  \end{center}

@@

\comment{
1. Nf3 b6 2. Nc3 g6 3. Ne5 g5 4. Nd5 b5
5. Nc4 bxc4 6. Nf4 gxf4 7. Rg1 c6 8. Rh1 c5
9. Rg1 f6 10. Rh1 f5 
}
\begin{center}
  \chessboard[setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11,
    pgfstyle=straightmove,markmoves={a6-b7,c3-b4,d6-c7,e6-f7,f3-g4,h6-g7},
  ]
\end{center}

The white b and g pawns have a clear route to promotion. White can
free each of the remaining 6 pawns with a single capture. Between this
and black's two pawn captures, this is the optimal 8 pawn moves
that are also captures. Since black has plenty of freed pieces, white
can promote all of their pawns during their own phase of critical
moves, resulting in:

\comment{
11. b3 Nc6 12. b4 Nd4
13. b5 Nb3 14. cxb3 Bb7 15. a3 Rc8 16. a4 Rc6
17. a5 Rd6 18. a6 Re6 19. axb7 Rd6 20. b8=N Re6
21. b6 Nh6 22. b7 Ng4 23. Nc6 Bh6 24. b8=Q Kf8
25. b4 Ke8 26. b5 Kf8 27. Qe5 Ke8 28. b6 Kf8
29. b7 Ke8 30. b8=N Bg5 31. h3 Bh4 32. hxg4 Bg3
33. fxg3 Qc8 34. g5 Qa6 35. Nb4 Kd8 36. Qe3 Kc7
37. g6 Kb6 38. g7 Rf6 39. g8=Q Re6 40. g4 Rf6
41. g5 Re6 42. g6 Rf6 43. g7 Re6 44. Qf8 Rf6
45. g8=N Rd6 46. Nf6 Qb5 47. Nh5 Rg8 48. g3 Rgg6
49. g4 Rge6 50. g5 Rf6 51. g6 Rc6 52. g7 Qa6
53. g8=N Qa5 54. d3 Qa6 55. d4 Qa5 56. d5 Qa6
57. dxc6 Qa5 58. c7 Qa6 59. c8=R Qa4 60. Qh3 Qa6
61. e3 Qa5 62. e4 Qa6 63. e5 Qa5 64. exf6 Qa6
65. Qg7 Qa5 66. f7 Qa6 67. Re8 Qa5 68. f8=R Qa6
69. Qdg4 Qa5 70. Nc6 Qa6 71. Ne5 Qa5 72. Nf3 Qa6
73. Nd3 Qa5+ 74. Ke2 Qa6 75. Nf2 Qa5 76. Qh8 Qa3
77. Nd1 Qa2+ 78. Nd2 Qa3 79. Kf2 Qa2 80. Kg1 Qa3
81. Rxa3
}
\begin{center}
  \chessboard[setfen=4RRNQ/p2pp2p/1k6/2p2p1N/2p2pQ1/R6Q/3N4/2BN1BKR b - - 0 81]
\end{center}

Now black can promote all of its pawns and capture white's pieces.
We actually leave the white queen; this turns out to be essential:

@@

\begin{center}
  \chessboard[setfen=1r6/8/1k3K2/8/2b5/2n4r/7Q/2r1q1rr w - - 0 133]
\end{center}

Finally, white captures all of black's pieces, and mates the black king.

@@

\begin{center}
  \chessboard[setfen=3k4/3Q4/4K3/8/8/8/8/8 b - - 2 145]
\end{center}

As we switch which color is making critical moves a total of three
times, we must come shy of the na\"ive maximum of 17,700 moves by
three. This gives us an upper bound of 17,697 for this approach,
which we will be able to achieve.

\paragraph{Ending condition.} \label{sec:ending} The way this game ends
is subtle for several reasons. First, note that we left the white
queen on the board and used it for mate. It is required that white be
the one mating for parity reasons, similar to the reason black must
make the first critical move. Since white makes critical moves in the
last phase, black leads on non-critical moves; at the moment white
makes the checkmating move, black has made 75 non-critical moves, and
white 74. White's 75th move mates.

@@

Why not capture the white queen too? This would be a critical move,
but once two kings remain, the game ends immediately in a draw. So
there is no length advantage here over checkmate. In fact we could do
worse: Consider the position in Figure~\ref{fig:insufficient} where
black is {\em forced} to capture the queen. This game is over prior to
the capture, so this is shorter than the checkmating sequence! Even if
the king had an escape square, it is arguable
(Section~\ref{sec:ambiguity}) that the game is over in any case: There
is no series of legal moves that leads to mate. Either the king
captures the queen (and then clearly no mate is possible with just the
two kings) or the king escapes, but in doing so plays the
75\th\ non-critical move, and triggers a draw by the 75-move rule.

@@

% XXX
% Are there any practical implications of the

@@

>>>>

@@

\title{Is this the longest Chess game?}
\author{Dr.~Tom~Murphy~VII~Ph.D.\thanks{
Copyright \copyright\ 2020 the Regents of the Wikiplia
Foundation. Appears in SIGBOVIK 2020 with the {\tt ShortMoveString} of the
Association for Computational Heresy; {\em IEEEEEE!} press,
Verlag-Verlag volume no.~0x40-2A.
\textcurrency 17,697
}
}

@@

In my experience, most chess games end in a few moves. If you want to
play a lot of chess moves, you just play a lot of chess games. Still,
there are games that seem to go on foreverrrrrrrrrr. Perhaps the
players are trying to {\em lull each other into a false sense of
security} while {\em waiting for the moment to strike}, or perhaps they
are stalling in a game of {\em Chess to the Death}.

Although many people ``know how to play'' chess, almost nobody fully
understands the rules of chess, most authoritatively given by
FIDE~\cite{fiderules}. (See for example Figure~\ref{fig:castle960} for
a minor chess scandal that erupted in 2019 over an obscure corner case
in the rules.) Several of these ``deep cuts'' have to do with
game-ending conditions that were introduced to avoid interminable
games.

@@

% This is good for the insurers and reinsurers of chess tournaments
% (games are guaranteed to end, rather than simply ending with
% probability 1).
So, chess is formally a finite game. This is good for computer
scientists, since it means that chess has a trivial O(1) optimal
solution. This allows us to move onto other important questions, like:
What is the longest chess game? In this paper I show how to compute
such a game, and then gratuitously present all of its 17,697 
moves. Even if you are a chess expert (``chexpert''), I bet you
will be surprised at some of the corner cases in the rules that are
involved.

@@

\begin{figure}[b!]
  \begin{center}
    \chessboard[
      % fillarea=a3-h1,printarea=a3-h1,
      marginbottom=false,margintop=false,
      setfen=2nr1r1k/pb1n1pqp/1p2p1p1/1N1pN3/5P2/1P4P1/P2PPQ1P/2R1K1RB,
      pgfstyle=curvemove,markmoves={e1-g1},
      pgfstyle=straightmove,markmoves={g1-f1},
      showmover=false]
  \end{center}
  \caption{ ({\it Nepomniachtchi -- So, 2019.}) White to move during a
    speed Chess960 (aka ``Fischer Random'') tournament. In this
    variant, the pieces start in different positions, but castling
    rules are such that the king and rook end up on the same squares
    that they would in normal chess. As a result, it is possible for
    the king or rook to not move during castling, or for the
    destination square for the king to already be occupied by the
    rook. Attempting to castle in the position depicted, grandmaster
    Ian Nepomniachtchi first touched the rook to move it out of the
    way. However, piece-touching rules require that when castling, the
    player first moves the king (and ``Each move must be played with
    one hand only''); but how? The rook is occupying g1! One commenter
    suggested tossing the king into the air, then sliding the rook to
    f1 while the king is airborne, and then watching the king land
    dead center on its target. The arbiter required Nepo to make a
    rook move instead, but this was later appealed, and the game
    replayed. } \label{fig:castle960}
\end{figure}

@@

\begin{quote}
\begin{itemize}
  \item[9.2.2.] Positions are considered the same if and only if the
same player has the move, pieces of the same kind and colour occupy
the same squares and the possible moves of all the pieces of both
players are the same. [...]
\comment{
  Thus positions are not the same if:
\begin{itemize}
\item[9.2.2.1.]	at the start of the sequence a pawn could have been captured en passant
\item[9.2.2.2.]	a king had castling rights with a rook that has not been moved, but forfeited these after moving. The castling rights are lost only after the king or rook is moved.
\end{itemize}
}
\end{itemize}
\end{quote}

@@

Fascinatingly (at least as fascinating as anything in this
questionable undertaking), this rule used to require {\em consecutive}
repetition of moves. However, there exist infinite sequences of moves
with no consecutive $n$-fold repetition. For example, in the starting
position, white and black can move either of their knights out and
back. Let {\bf 0} be \knight c3 \Knight c6 \knight b1 \Knight b8
(queenside knights move, returning to the starting position) and {\bf
  1} be \knight f3 \Knight f6 \knight g1 \Knight g8 (kingside). Now
the Prouhet--Thue--Morse sequence~\cite{thue1906uber}
0110100110010110\ldots\footnote{
  Let $s_0$ be the string {\tt "0"}, then
  $s_i$ is $s_{i-1} \overline{s_{i-1}}$
  where $\overline{\tt 0} = {\tt 1}$ and $\overline{\tt 1} = {\tt 0}$.
}
%
can be executed. This infinite
sequence is cube-free (does not contain $SSS$ for any non-empty string
$S$)~\cite{oeisa010060}, and therefore never violates the consecutive
threefold repetition rule~\cite{euwe1929mengentheoretische}.

@@

\begin{figure}[htp]
  \begin{center}
    \chessboard[margintop=false,marginbottom=false,setfen=8/8/5K2/8/8/8/Q7/k7 b - - 0 1]
  \end{center}
  \caption{Black to move and draw in 0 (!). Most players and even
    chexperts believe that the only legal move is Kxa2, and
    that the game then ends in a draw with ``insufficient material.''
    However, this game is {\em already over}. Since neither black
    nor white can win via any series of legal moves, by rule 5.2.2
    the game immediately ends in a `dead position'. (Although see
    Section~\ref{sec:ambiguity} for possible ambiguity in this
    rule.)} \label{fig:insufficient}
\end{figure}

@@

\begin{figure}[htp]
  \begin{center}
    \chessboard[margintop=false,marginbottom=false,setfen=k1bB4/8/2p1p1p1/1pPpPpP1/1P1P1P2/8/8/K1Bb4 w - - 0 1]
  \end{center}
  \caption{Thinking of implementing the rules of chess? To be correct,
    you'll need your program to be able to deduce that no helpmates are
    possible in this position and thus the game is over. Stockfish rates
    this as +0.4 for white, even searching to depth 92.
    (Position is due to user {\tt supercat} on Chess StackExchange.)
  } \label{fig:nohelp}
\end{figure}

@@

First, the rules specifically define ``legal move'', with 3.10.1
saying ``A move is {\em legal} when all the relevant requirements of
Articles 3.1 -- 3.9 have been fulfilled.'' These requirements describe
the movement of each piece as you are familiar (e.g. 3.3 ``The rook
may move to any square along the file or the rank on which it
stands.''). They also disallow capturing one's own pieces, or moving
when in check. However, they allow as legal some moves that would
otherwise be prohibited, like capturing the opponent's king (this is
excluded by 1.4.1, outside the definition of ``legal''). Capturing the
opponent's king is generally not useful for demonstrating that
checkmate is possible (Figure~\ref{fig:kingcapture}), so this is
mostly a curiosity.

\begin{figure}
  \begin{center}
    \begin{tabular}{cc}
      \chessboard[smallboard,marginleft=false,marginright=false,setfen=BBKBRRnk/PPPPPPnn/3P1BRR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] &
      \chessboard[smallboard,marginright=false,setfen=BBKBRR1B/PPPPnP1n/3P2RR/3PBPPR/3P1P1P/8/8/8 w - - 0 1] \\
      (a) & (b) \\
  \end{tabular}
  \end{center}

@@

\comment{
1. Nf3 b6 2. Nc3 g6 3. Ne5 g5 4. Nd5 b5
5. Nc4 bxc4 6. Nf4 gxf4 7. Rg1 c6 8. Rh1 c5
9. Rg1 f6 10. Rh1 f5 
}
\begin{center}
  \chessboard[
    margintop=false,marginbottom=false,
    setfen=rnbqkbnr/p2pp2p/8/2p2p2/2p2p2/8/PPPPPPPP/R1BQKB1R w Qkq - 0 11,
    pgfstyle=straightmove,markmoves={a6-b7,c3-b4,d6-c7,e6-f7,f3-g4,h6-g7},
  ]
\end{center}

\smallskip
\noindent The white b and g pawns have a clear route to promotion. White can
free each of the remaining 6 pawns with a single capture. Between this
and black's two pawn captures, this is the optimal 8 pawn moves
that are also captures. Since black has plenty of freed pieces, white
can promote all of their pawns during their own phase of critical
moves, resulting in:

\comment{
11. b3 Nc6 12. b4 Nd4
13. b5 Nb3 14. cxb3 Bb7 15. a3 Rc8 16. a4 Rc6
17. a5 Rd6 18. a6 Re6 19. axb7 Rd6 20. b8=N Re6
21. b6 Nh6 22. b7 Ng4 23. Nc6 Bh6 24. b8=Q Kf8
25. b4 Ke8 26. b5 Kf8 27. Qe5 Ke8 28. b6 Kf8
29. b7 Ke8 30. b8=N Bg5 31. h3 Bh4 32. hxg4 Bg3
33. fxg3 Qc8 34. g5 Qa6 35. Nb4 Kd8 36. Qe3 Kc7
37. g6 Kb6 38. g7 Rf6 39. g8=Q Re6 40. g4 Rf6
41. g5 Re6 42. g6 Rf6 43. g7 Re6 44. Qf8 Rf6
45. g8=N Rd6 46. Nf6 Qb5 47. Nh5 Rg8 48. g3 Rgg6
49. g4 Rge6 50. g5 Rf6 51. g6 Rc6 52. g7 Qa6
53. g8=N Qa5 54. d3 Qa6 55. d4 Qa5 56. d5 Qa6
57. dxc6 Qa5 58. c7 Qa6 59. c8=R Qa4 60. Qh3 Qa6
61. e3 Qa5 62. e4 Qa6 63. e5 Qa5 64. exf6 Qa6
65. Qg7 Qa5 66. f7 Qa6 67. Re8 Qa5 68. f8=R Qa6
69. Qdg4 Qa5 70. Nc6 Qa6 71. Ne5 Qa5 72. Nf3 Qa6
73. Nd3 Qa5+ 74. Ke2 Qa6 75. Nf2 Qa5 76. Qh8 Qa3
77. Nd1 Qa2+ 78. Nd2 Qa3 79. Kf2 Qa2 80. Kg1 Qa3
81. Rxa3
}
\begin{center}
  \chessboard[
    margintop=false,marginbottom=false,
    setfen=4RRNQ/p2pp2p/1k6/2p2p1N/2p2pQ1/R6Q/3N4/2BN1BKR b - - 0 81]
\end{center}

\smallskip
\noindent Now black can promote all of its pawns and capture white's pieces.
We actually leave the white queen; this turns out to be essential:

@@

\begin{center}
  \chessboard[
    margintop=false,marginbottom=false,
    setfen=1r6/8/1k3K2/8/2b5/2n4r/7Q/2r1q1rr w - - 0 133]
\end{center}

\smallskip
\noindent Finally, white captures all of black's pieces, and mates the black king:

@@

\begin{center}
  \chessboard[
    margintop=false,marginbottom=false,
    setfen=3k4/3Q4/4K3/8/8/8/8/8 b - - 2 145]
\end{center}

\smallskip
\noindent Since we switch which color is making critical moves a total of three
times, we must come shy of the na\"ive maximum of 17,700 moves by
three. This gives us an upper bound of 17,697 for this approach,
which we will be able to achieve.

\paragraph{Ending condition.} \label{sec:ending} The way this game ends
is subtle for several reasons. First, note that we left the white
queen on the board and used it for mate. It is required that white be
the one mating for parity reasons, similar to the reason black must
make the first critical move. Since white makes critical moves in the
last phase, black leads on non-critical moves; at the moment white
makes the checkmating move, black has made 75 non-critical moves, and
white 74. White's 75\th\ move mates.

@@

Why not capture the white queen too? This would be a critical move,
but once two kings remain, the game ends immediately in a draw. So
there is no length advantage here over checkmate. In fact, attempting
to capture may {\em shorten} the game: Consider the position in
Figure~\ref{fig:insufficient} where black is {\em forced} to capture
the queen. This game is over prior to the capture, so this is shorter
than the checkmating sequence! Even if the king had an escape square,
it is arguable (Section~\ref{sec:ambiguity}) that the game is over in
any case: There is no ``series of legal moves'' that leads to mate.
Either the king captures the queen (and then clearly no mate is
possible with just the two kings) or the king escapes, but in doing so
plays the 75\th\ non-critical move, and triggers a draw by the 75-move
rule.

@@

@@

\smallskip
The character called \queen\ often plays an important role in the story.
How would you describe her personality? How does she develop over the
course of the game?
